Language learning via recursive pragmatic reasoning

### 2013-05-31 praglearn.tex
merge

Natural language is a powerful tool for conveying meaning in
context. Consider the utterances
``it's raining,'' ``I ate some of the cookies,'' or ``can you close
the window?'' In each, a listener must go beyond the literal meaning
of the words to fill in contextual details (``it's raining here and
now''), infer that a stronger alternative is not true (``I ate some
but not all of the cookies''), or more generally infer the speaker's
communicative goal (``I want you to close the window right now because
I'm cold''), a process known as \textit{pragmatic reasoning}. Theories of pragmatics frame
the process of language comprehension as inference about the
generating goal of an utterance given a rational speaker
\cite{grice1975,dale1995,frank2012}. For example, a listener might
reason, ``if she had wanted me to think `all' of the cookies, she
would have said `all'---but she didn't. Hence `all' must not be true
and she must have eaten some {\it but not all} of the cookies.'' This kind of reasoning is core to language use.

But pragmatic reasoning often intersects with language learning as
well. In both adults and children, uncertainty about word meanings is
common, and often considering speakers' pragmatic goals can help to resolve this uncertainty. For example,
if a novel word is used in a context containing both a novel and a
familiar object, young children can make the inference
that the novel word refers to the novel object \cite{markman1988}.\footnote{Very young children
  make inferences that are often labeled as ``pragmatic'' in that they
  involve reasoning about context \cite{clark1988,baldwin1993}, though
  they also show weaknesses in specific inferences (e.g.~strengthening
  {\sc some} to {\sc some but not all} \cite{papafragou2003}). Here we
  remain agnostic about the age at which children are able to make
  such inferences robustly, as it may vary depending on the linguistic
  materials being used in the inference \cite{barner2011}.} For adults
who are proficient language users, there are also a variety of intriguing cases
in which listeners seem to create situation- and task-specific ways of
referring to particular objects. For example, when asked to refer to
idiosyncratic geometric shapes, over the course of an experimental
session, participants create conventionalized descriptions that allow
them to perform accurately even though they do not begin with a shared
label \cite{krauss1964,clark1986}. In both of these examples, reasoning about another person's goals informs language learners' estimates of what words are likely to mean.

@@

\begin{table}[t]
\label{tab:results}
\begin{center}
\begin{tabular}{lccccc}
\toprule
Phenomenon & Ref. & WL & PI & PI+U & PI+WL \\
\midrule
Interpreting specificity implicature & \cite{grice1975} &  & x &x & x\\
Interpreting Horn implicature & \cite{horn1984} & &  & x& x\\
Cross-situational learning & \cite{smith2008} & x& & & x \\
Disambiguating new words using old words & \cite{markman1988} &x  & & & x \\
Disambiguation without mapping & \cite{horst2008} &  & & & x \\
Emergence of novel \& efficient communicative systems & \cite{galantucci2005} & & & & x \\
Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
Non-lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
\hline
\end{tabular}
\end{center}
\caption{Empirical results and references. WL refers to the word
  learning model of \cite{frank2009}; PI refers to the recursive
  pragmatic inference model of \cite{frank2012}; PI+U refers to the
  pragmatic inference model of \cite{bergen2012} which includes
  lexical uncertainty, marginalizes it out, and then recurses. Our
  current model is referred to here as PI+WL, and combines pragmatic
  inference with word learning.}
\end{table}

@@

% agents not sure whether some literally means some-not-all or not,
% could lexicalize given the right kind of input, but regular pragmatic
% strengthening alone is not enough

>>>>

Two puzzles present themselves to language users: What do words mean in general and what do they mean in context?
%Natural language is a powerful tool for conveying meaning in context. 
Consider the utterances
``it's raining,'' ``I ate some of the cookies,'' or ``can you close
the window?'' In each, a listener must go beyond the literal meaning
of the words to fill in contextual details (``it's raining here and
now''), infer that a stronger alternative is not true (``I ate some
but not all of the cookies''), or more generally infer the speaker's
communicative goal (``I want you to close the window right now because
I'm cold''), a process known as \textit{pragmatic reasoning}. Theories of pragmatics frame
the process of language comprehension as inference about the
generating goal of an utterance given a rational speaker
\cite{grice1975,dale1995,frank2012}. For example, a listener might
reason, ``if she had wanted me to think `all' of the cookies, she
would have said `all'---but she didn't. Hence `all' must not be true
and she must have eaten some {\it but not all} of the cookies.'' This kind of reasoning is core to language use.

But pragmatic reasoning about meaning-in-context relies on stable literal meanings that must  themselves be learned.
%and often intersects with learning the meaning of language in general. 
In both adults and children, uncertainty about word meanings is
common, and often considering speakers' pragmatic goals can help to resolve this uncertainty. For example,
if a novel word is used in a context containing both a novel and a
familiar object, young children can make the inference
that the novel word refers to the novel object \cite{markman1988}.\footnote{Very young children
  make inferences that are often labeled as ``pragmatic'' in that they
  involve reasoning about context \cite{clark1988,baldwin1993}, though
  they also show weaknesses in specific inferences (e.g.~strengthening
  {\sc some} to {\sc some but not all} \cite{papafragou2003}). Here we
  remain agnostic about the age at which children are able to make
  such inferences robustly, as it may vary depending on the linguistic
  materials being used in the inference \cite{barner2011}.} For adults
who are proficient language users, there are also a variety of intriguing cases
in which listeners seem to create situation- and task-specific ways of
referring to particular objects. For example, when asked to refer to
idiosyncratic geometric shapes, over the course of an experimental
session, participants create conventionalized descriptions that allow
them to perform accurately even though they do not begin with a shared
label \cite{krauss1964,clark1986}. In both of these examples, reasoning about another person's goals informs language learners' estimates of what words are likely to mean.

@@

\begin{table}[t]
\label{tab:results}
\begin{center}
\begin{tabular}{lccccc}
\toprule
Phenomenon & Ref. & WL & PI & PI+U & PI+WL \\
\midrule
Interpreting specificity implicature & \cite{grice1975} &  & x &x & x\\
Interpreting Horn implicature & \cite{horn1984} & &  & x& x\\
Cross-situational learning & \cite{smith2008} & x& & & x \\
Disambiguating new words using old words & \cite{markman1988} &x  & & & x \\
Disambiguation without mapping & \cite{horst2008} &  & & & x \\
Learning/lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
Emergence of novel \& efficient communicative systems & \cite{galantucci2005} & & & & x \\
Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
Non-lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
\hline
\end{tabular}
\end{center}
\caption{Empirical results and references. WL refers to the word
  learning model of \cite{frank2009}; PI refers to the recursive
  pragmatic inference model of \cite{frank2012}; PI+U refers to the
  pragmatic inference model of \cite{bergen2012} which includes
  lexical uncertainty, marginalizes it out, and then recurses. Our
  current model is referred to here as PI+WL, and combines pragmatic
  inference with word learning.}
\end{table}

@@

agents not sure whether some literally means some-not-all or not,
could lexicalize given the right kind of input, but regular pragmatic
strengthening alone is not enough
